Automatically proving equivalence by type-safe reflection

### 2016-01-13 paper-JFP-DTP/content/2-basicIdeas.tex
paper continued, section 1&2, typos...

We will first present the basic ideas of our technique on a simplified version, in which we aim to deal with universally quantified natural numbers, the properties of associativity and neutral element for the operation plus.
For example, we would like to be able to automatically generate proofs of goals like $\forall x1\ x2\ x3:Nat,\ (x1 + l2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$ [Example 1]. \\
For this smaller problem, we have decided to only work with the associativity of plus : $\forall x1\ x2\ x3,\ (x1 + x2) + x3 = x1 + (x2 + x3)$ and with the fact that Z is a right neutral element for the $+$ operation : $\forall x, x + Z = x$. Note that $Z$ is also a left neutral element, but this property is not needed, as we have this behaviour by reduction as $+$ is defined recursively on its first argument. Of course, some structures can have more or less properties than these two, and this work will be extended in the next sections when we will write a general hierarchy of solvers for different algebraic structures. \\
Thus, in definitive, in this section, we want to write a decision procedure, able to tell if two expressions composed of universally quantified natural numbers and additions of these numbers are equal, and to produce a proof of this equality if possible, when "equal" has the meaning "syntactically equal or equal thanks to the associative and neutral properties".

@@

When trying to prove this kind of equalities, the variables are abstracted, and they become part of the context. On the example 1, after abstraction of the variables, the goal becomes simply $(x1 + x2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$, which is something of the general form $x=y$.
The general idea --that will also apply for the more general and bigger problem detailed in the next sections-- will be to normalize both sides of the "potential equality" $x=y$, and afterwards to compare them using Leibniz syntactical equality.
The goal of the normalization is to compute a canonical representation for a number $x$, such that any other number provably equal to $x$ (by using the two available properties) will have the same canonical representation. For example, the normalisation might transform $x+((y+Z)+z)$ into $(x+y)+z$ if we decide that the normal form will be completely left associative, and every addition between an element $a$ and zero should be simplified to $a$. It will then be possible to decide the equality by simply comparing the normalised left and right hand sides with a simple syntactical equality. This is in fact what we do all the time that we have to decide if two things, written differently, are equal or not. When a human is given two mathematical polynomials and has to decide the equality of these functions, a technique that always works is to decide once and for all a canonical representation of polynomials, and to put both ponynomials in this form. If the normalised form are the same, then the two original polynomials are equal, otherwise they do not represent the same computation.\\
\\
In fact, such a normalisation function can't be written directly, because in the LHS and RHS of $x=y$, we potentially have variables which have been universally quantified. And the normalization function needs to do different treatments for a "variable natural number" (a number which has been universally quantified) and for the constant $Z$. This is not possible yet, because once the variables are abstracted, they are just ordinary values of type $Nat$, and there is no way to distinguish them. Indeed, this information only exists at the level of the ASTs representing the two terms, and we do not have a direct access to these syntax trees.
For this reason, we will work by reflection. In this little example, it means that we will define a data type which will be used as an encoding of natural numbers, or more precisely, an encoding of natural numbers composed of "variable numbers", $Z$, and additions of these things. This data type will allow us to know the internal structure of a number, ie, where the variables and constants are.
Indeed, this data type will allow pattern matching. Previously, we were only able to pattern match a natural number against the constructors $Z$ and $S$, which wasn't what we needed.  With the first approximation of the data type $Expr$ presented in the Fig 6, we will be allowed to pattern match an encoding of number against the constructor $Plus$, $Var$ and $Zero$, which gives us the informations we want.

@@

If we continue with this first definition of $Expr$, the normalisation function will certainly take an Expr and produce another Expr, and we will need to prove the following lemma afterwards : \\
$\forall\ e:Expr\ \Gamma,\ reify\ (reduce\ e)\ =\ reify\ e$ \\
where $reify$ is a function computing the interpretation of an Expr in a context $\Gamma$, that is to say, the natural number that this Expr is encoding.
This proof can be quite tricky to build because it relies on the complete behaviour of the reduction and on the way the interpretation is computed.
For this little example, the reduction procedure will not be too heavy, but in the next sections, with more sophisticated algebraic structures, we will have more rewriting rules to deal with and it will certainly become more problematic to "unfold" the definition of a gigantic reduction procedure. As much as possible, we want to avoid the pain of proving the correctness of the tactic, and for this reason, we don't want to have to write huge proofs. \\
\\
To avoid this source of complexity, we add an index to the type $Expr$, and this index is precisely the concrete number that the Expr is encoding. This is the first brick to our type-safe reflection mechanism. Thus, it won't be necessary to define the $reify$ function, as we will know directly the concrete element reflected by a term of type $Expr\ \Gamma\ x$ just by looking at its index $x$. Something more important is that we will also have a guarantee that the reflection of an expression $e$ is indeed a faithful representation of $e$ \\

@@

We want to write the reduction function on a "correct by construction" way, which means that no additional proof should be required after the definition of the function. Thus, $reduce$ will produce the proof that the new Expr freshly produced has the same interpretation as the original Expr, and this will be made easier by the fact that the data type Expr is now indexed over the real --concrete-- number : a term of type $Expr\ \Gamma\ x$ is the encoding of the number $x$.
Thus, we can write the type of $reduce$ like this : \\
$reduce\ :\ Expr\ \Gamma\ x\ \rightarrow\ (x'\ **\ (Expr\ \Gamma\ x',\ x\ =\ x'))$ \\
The function $reduce$ produces a dependent pair : the new concrete number $x'$, and a pair made of an $Expr\ \Gamma\ x'$ which is the new encoded term indexed over the new concrete number we have just produced, and a proof that old and new --concrete-- numbers are equal.
Note that this function can't simply produce an $Expr\ \Gamma\ x$, because the number on which the resulting expression will be indexed is not necessary syntactically equal to the original number since this equality can use the two available properties. Said differently, even if we can prove $x=x'$ (if the function is correctly defined), we do not have $x \equiv x'$.
And in fact, what really interest us in this function is precisely the proof of $x\ =\ x'$.
The reason is that when we try to automatically prove $x=y$, these proofs $x=x'$ and $y=y'$ will be the crucial part for the construction of the desired proof. \\
\\
We have an expression $ex$ encoding $x$, and an expression $ey$ encoding $y$. We will normalize $ex$, and this will give a new concrete number $x'$, a new expression $ex':Expr\ \Gamma\ x'$, and a proof of $x=x'$. We will do the same with $ey$, and we will get a new concrete number $y'$, an expression $ey':Expr\ \Gamma\ y'$, and a proof of $y=y'$. \\
It is now enough to simply compare $ex'$ and $ey'$ using a standard syntactical equality because these two expressions are now supposed to be in normal form :

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
eqExpr : (e : Expr gam x) -> (e' : Expr gam y) -> Maybe (e = e')
eqExpr (Plus x y) (Plus x' y') with (eqExpr x x', eqExpr y y')
  eqExpr (Plus x y) (Plus x y) | (Just refl, Just refl) = Just refl
  eqExpr (Plus x y) (Plus x' y') | _ = Nothing
eqExpr (Var i) (Var j) with (decEq i j)
  eqExpr (Var i) (Var i) | (Yes refl) = Just refl
  eqExpr (Var i) (Var j) | _ = Nothing
eqExpr Zero Zero = Just refl
eqExpr _ _ = Nothing
\end{verbatim}
\end{center}
\caption{Syntactical equality between reflected terms}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
buildProof : {x : Nat} -> {y : Nat} -> Expr gam x' -> Expr gam y' 
           -> (x = x') -> (y = y') -> Maybe (x = y)
buildProof ex' ey' lp rp with (eqExpr ex' ey')
  buildProof ex' ex' lp rp | Just refl = ?MbuildProof
  buildProof ex' ey' lp rp | Nothing = Nothing
\end{verbatim}
\end{center}
\caption{Building the desired proof with the two proofs of equality}
\figrule
\end{figure}

The $Expr\ \Gamma\ x'$ is the normalised reflected left hand size of the equality, which represents the value $x'$. Before the normalisation, the reflected LHS was reflecting the value $x$. The $Expr\ \Gamma\ y'$ is the normalised reflected right hand size, which now represents the value $y'$, but which was encoding $y$ before the normalisation. The function also expects the proofs that $x=x'$ and $y=y'$, and we should be able to pass them because the normalisation function also produces the proof of equality between the previous and the new concrete values.

As we mentioned, the proof for the metavariable $MbuildProof$ is just a rewriting of the two equalities :

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
  MbuildProof = proof {
  intros; refine Just; rewrite sym p1; rewrite sym p2; exact refl;
}  
\end{verbatim}
\end{center}
\caption{buildProof metavariable}
\figrule
\end{figure}

Finally, the main function which tries to prove the equality $x=y$ simply has to reduce the two metaified terms reflecting the left and the right hand side, and to use the function $buildProof$ in order to compose the two proofs we just obtained :

@@

Now, we need to define the function reduce. To do that, we have to decide a canonical representation of associative natural numbers. We decide that the left associative form will be the canonical representation. Thus, the $reduce$ function has to rewrite the reflected term by rearranging the parentheses in order to transform the underlying number in the form $(...((x1 + x2) + x3) ... + xn)$. To do so, one possibility is to define a new datatype which captures this property, and to write a function going from $Expr$ to this new type. Thus it will be easier to be certain that we are effectively computing the normal form : forcing properties to hold by the shape of a datatype is a good usage of dependent types when, like here, it doesn't introduce more complications.

@@

As part of the normalization, we write a function $expr\_l$ which converts an $Expr\ \Gamma\ x$ to a $LExpr\ \Gamma\ x'$ and which produces a proof of $x=x'$. This function will therefore use the two available properties multiple times --and especially the property of associativity--, in order to obtain the expected fully left associative form.

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
expr_l : Expr gam x 
         -> (x' ** (LExpr gam x', x = x'))
expr_l Zero = (_ ** (LZero, refl))
expr_l (Var i) = (_ ** (LPlus LZero i, refl))
expr_l (Plus ex ey) = 
  let (xl ** (xr, xprf)) = expr_l ex in
  let (yl ** (yr, yprf)) = expr_l ey in
    plusLExpr _ _ xr yr xprf yprf
      where 
      plusLExpr : (x', y' : Nat)
            -> {gam : Vect n Nat} -> {x, y : Nat} 
            -> LExpr gam x -> LExpr gam y 
            -> (x' = x) -> (y' = y) 
            -> (w' ** (LExpr gam w', x'+y'=w'))
      plusLExpr x' y' rx (LPlus e i) xprf yprf =
        let (xrec ** (rec, prf)) = 
          plusLExpr _ _ rx e refl refl in
          (_ ** (LPlus rec i, ?plusLExpr1))
      plusLExpr x' y' rx LZero xprf yprf =
        (_ ** (rx, ?plusLExpr2))
\end{verbatim}
\end{center}
\caption{Production of the left associative form}
\figrule
\end{figure}

Using this new datatype $LExpr$ has changed the representation of our encoded lists, so we need to convert back an $LExpr\ \Gamma\ x$ to an $Expr\ \Gamma\ x$. The function $l\_expr$ does this easy task.
\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
l_expr : LExpr gam x -> Expr gam x
l_expr LZero = Zero
l_expr (LPlus x i) = Plus (l_expr x) (Var i)
\end{verbatim}
\end{center}
\caption{Going back from LExpr to Expr}
\figrule
\end{figure}

We notice that in order to transform the expression into its left associative equivalent representation, we've effectively needed to know where the variables and the $Z$ constants are : the functions $expr\_l$ and $l\_expr$ are doing different treatments for these different possibilities. \\
\\
We've got two metavariables to prove. The metavariable $plusLExpr1$ requires us to prove the goal : $x' + y' = xrec + index\ i\ \Gamma$ in a context where we've got, amongst other things,  $(xprf\ :\ x'\ =\ x)$, $(yprf\ :\ y'\ =\ x1\ +\ index\ i\ \Gamma)$ and $(prf\ :\ x\ +\ x1\ =\ xrec)$.
Proving this goal uses the property of associativity after rewriting the goal with these three proof of equality $xprf$, $yprf$ and $prf$.

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
plusLExpr1 = proof {
  intros; rewrite sym xprf; rewrite sym yprf; rewrite prf;
  mrefine plusAssociative;
}
\end{verbatim}
\end{center}
\caption{Proof of the metavariable plusLExpr1}
\figrule
\end{figure}

And the metavariable plusLExpr2 uses the fact that $Z$ is a neutral element for the addition.

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
plusLExpr2 = proof
  intros; rewrite xprf; rewrite (sym yprf); mrefine plusZeroRightNeutral
\end{verbatim}
\end{center}
\caption{Proof of the metavariable plusLExpr2}
\figrule
\end{figure}

>>>>

We will first present the basic ideas of our technique on a simplified problem, in which we aim to deal with universally quantified natural numbers and additions.
For example, we would like to be able to automatically generate proofs of goals like $\forall x1\ x2\ x3:Nat,\ (x1 + l2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$ [Example 1]. \\
For this smaller problem, we have decided to only work with the associativity of $+$ : $plusAssociative : \forall x1\ x2\ x3,\ (x1 + x2) + x3 = x1 + (x2 + x3)$ and with the fact that Z is a right neutral element for the $+$ operation : $plusZeroRightNeutral : \forall x, x + Z = x$. Note that $Z$ is also a left neutral element, but this property is not needed because we have this behaviour by reduction, as $+$ is defined recursively on its first argument. Of course, some structures can have more or less properties than these two, and this work will be extended in the next sections when we will write a general hierarchy of provers for different algebraic structures. \\
Thus, in definitive, in this section, we want to write a decision procedure, able to tell if two expressions composed of universally quantified natural numbers and additions of these numbers are equal, and to produce a proof of this equality if possible, when "equal" has the meaning "syntactically equal or equal thanks to the associative and neutral properties".

@@

When trying to prove this kind of equalities, the variables are abstracted, and they become part of the context. On the example 1, after abstraction of the variables, the goal becomes simply $(x1 + x2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$, which is something of the general form $x=y$.
The general idea --that will also apply for the more general and bigger problem detailed in the next sections-- will be to normalize both sides of the "potential equality" $x=y$, and afterwards to compare them using Leibniz syntactical equality.
The goal of the normalization is to compute a canonical representation for a number $x$, such that any other number provably equal to $x$ (by using the two available properties) will have the same canonical representation. For example, the normalisation might transform $x+((y+Z)+z)$ into $(x+y)+z$ if we decide that the normal form will be completely left associative, and every addition between an element $a$ and zero should be simplified to $a$. It will then be possible to decide the equality by simply comparing the normalised left and right hand sides with a simple syntactical equality. This is in fact what we do all the time that we have to decide if two things, written differently, are equal or not. When a human is given two mathematical polynomials and has to decide the equality of these functions, a technique that always works is to decide once and for all a canonical representation of polynomials, and to put both polynomials in this form. If the normalised forms are the same, then the two original polynomials are equal, otherwise they do not represent the same computation.\\
\\
In fact, such a normalisation function can't be written directly, because in the LHS and RHS of $x=y$, we potentially have variables which have been universally quantified. And the normalization function needs to do different treatments for a "variable natural number" (a number which has been universally quantified) and for the constant $Z$. This is not possible yet, because once the variables are abstracted, they are just ordinary values of type $Nat$, and there is no way to distinguish them. Indeed, this information only exists at the level of the ASTs representing the two terms, and we do not have a direct access to these syntax trees.
For this reason, we will work by reflection. In this little example, it means that we will define a data type which will be used as an encoding of natural numbers, or more precisely, an encoding of natural numbers composed of "variable numbers", $Z$, and additions of these things. This data type will allow us to know the internal structure of a number, ie, where the variables and constants are.
Indeed, this data type will allow pattern matching. Previously, we were only able to pattern match a natural number against the constructors $Z$ and $S$, which wasn't what we needed.  With the first approximation of the data type $Expr$ presented in the Fig 6, we will be allowed to pattern match an encoding of number against the constructor $Plus$, $Var$ and $Zero$, which gives us the informations we want.

@@

If we continue with this first definition of $Expr$, the normalisation function will certainly take an Expr and produce another Expr, and we will need to prove the following lemma afterwards : \\
$\forall\ e:Expr\ \Gamma,\ reify\ (reduce\ e)\ =\ reify\ e$ \\
where $reify$ is a function computing the interpretation of an Expr in a context $\Gamma$, that is to say, the natural number that this Expr is encoding.
This proof can be quite tricky to build because it relies on the complete behaviour of the reduction and on the way the interpretation is computed.
For this little example, the reduction procedure will not be too heavy, but in the next sections, with more sophisticated algebraic structures, we will have more rewriting rules to deal with and it will certainly become more problematic to "unfold" the definition of a gigantic reduction procedure. We want to avoid pain as much as possible, and for this reason, we don't want to have to write huge proofs. \\
\\
To avoid this source of complexity, we add an index to the type $Expr$, and this index is precisely the concrete number that the Expr is encoding. This is the first brick to our type-safe reflection mechanism. Thus, it won't be necessary to define the $reify$ function, as we will know directly the concrete element reflected by a term of type $Expr\ \Gamma\ x$ just by looking at its index $x$. Something more important is that we will also have a guarantee that the reflection of an expression $e$ is indeed a faithful representation of $e$ \\

@@

We want to write the reduction function on a "correct by construction" way, which means that no additional proof should be required after the definition of the function. Thus, $reduce$ will produce the proof that the new Expr freshly produced has the same interpretation as the original Expr, and this will be made easier by the fact that the data type Expr is now indexed over the real --concrete-- number : a term of type $Expr\ \Gamma\ x$ is the encoding of the number $x$.
Thus, we can write the type of $reduce$ like this : \\
$reduce\ :\ Expr\ \Gamma\ x\ \rightarrow\ (x'\ **\ (Expr\ \Gamma\ x',\ x\ =\ x'))$ \\
The function $reduce$ produces a dependent pair : the new concrete number $x'$, and a pair made of an $Expr\ \Gamma\ x'$ which is the new encoded term indexed over the new concrete number we have just produced, and a proof that old and new --concrete-- numbers are equal.
Note that this function can't simply produce an $Expr\ \Gamma\ x$, because the concrete number on which the resulting expression will be indexed is not necessary syntactically equal to the original number since this equality can use the two available properties. Said differently, even if we can prove $x=x'$ (if the function is correctly defined), we do not have $x \equiv x'$.
And in fact, what really interests us in this function is precisely this proof of $x\ =\ x'$.
The reason is that when we try to automatically prove $x=y$, these proofs will be the crucial part for the construction of the desired proof. \\
\\
The tactic will work as follow.
We have an expression $ex$ encoding $x$, and an expression $ey$ encoding $y$. We will normalize $ex$, and this will give a new concrete number $x'$, a new expression $ex':Expr\ \Gamma\ x'$, and a proof of $x=x'$. We will do the same with $ey$, and we will get a new concrete number $y'$, an expression $ey':Expr\ \Gamma\ y'$, and a proof of $y=y'$. \\
It is now enough to simply compare $ex'$ and $ey'$ using a standard syntactical equality because these two expressions are now supposed to be in normal form :

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
eqExpr : (e : Expr gam x) -> (e' : Expr gam y) -> Maybe (e = e')
eqExpr (Plus x y) (Plus x' y') with (eqExpr x x', eqExpr y y')
  eqExpr (Plus x y) (Plus x y) | (Just Refl, Just Refl) = Just Refl
  eqExpr (Plus x y) (Plus x' y') | _ = Nothing
eqExpr (Var i) (Var j) with (decEq i j)
  eqExpr (Var i) (Var i) | (Yes Refl) = Just Refl
  eqExpr (Var i) (Var j) | _ = Nothing
eqExpr Zero Zero = Just Refl
eqExpr _ _ = Nothing
\end{verbatim}
\end{center}
\caption{Syntactical equality between reflected terms}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
buildProof : {x : Nat} -> {y : Nat} -> Expr gam x' -> Expr gam y' 
           -> (x = x') -> (y = y') -> Maybe (x = y)
buildProof ex' ey' lp rp with (eqExpr ex' ey')
  buildProof ex' ex' lp rp | Just Refl = ?MbuildProof
  buildProof ex' ey' lp rp | Nothing = Nothing
\end{verbatim}
\end{center}
\caption{Building the desired proof with the two proofs of equality}
\figrule
\end{figure}

The argument of type $Expr\ \Gamma\ x'$ is the normalised reflected left hand size of the equality, which represents the value $x'$. Before the normalisation, the reflected LHS was reflecting the value $x$. The $Expr\ \Gamma\ y'$ is the normalised reflected right hand size, which now represents the value $y'$, but which was encoding $y$ before the normalisation. The function also expects the proofs that $x=x'$ and $y=y'$, and we should be able to pass them because the normalisation function has also produced the proof of equality between the original and the new concrete values.

As mentioned, the proof for the metavariable $MbuildProof$ is just a rewriting of the two equalities :

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
  MbuildProof = proof {
  intros; refine Just; rewrite sym p1; rewrite sym p2; exact Refl;
}  
\end{verbatim}
\end{center}
\caption{buildProof metavariable}
\figrule
\end{figure}

Finally, the main function which tries to prove the equality $x=y$ simply has to reduce the two reflected terms encoding the left and the right hand side, and to use the function $buildProof$ in order to compose the two proofs that we just obtained :

@@

We now need to define the function reduce. To do that, we have to decide a canonical representation of associative natural numbers. We decide that the left associative form will be the canonical representation. Thus, the $reduce$ function has to rewrite the reflected terms by rearranging the parentheses in order to transform the underlying concrete number in the form $(...((x1 + x2) + x3) ... + xn)$. To do so, one possibility is to define a new datatype which captures this property, and to write a function going from $Expr$ to this new type. Thus it will be easier to be certain that we are effectively computing the normal form : forcing properties to hold by the shape of a datatype is a good usage of dependent types when, like here, it doesn't introduce more complications.

@@

As part of the normalization, we write a function $expr\_l$ which converts an $Expr\ \Gamma\ x$ to a $LExpr\ \Gamma\ x'$ and which produces a proof of $x=x'$. This function will therefore use the two available properties multiple times, while rewriting the term until the fully left associative desired form is obtained.

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
expr_l : Expr gam x 
         -> (x' ** (LExpr gam x', x = x'))
expr_l Zero = (_ ** (LZero, Refl))
expr_l (Var i) = (_ ** (LPlus LZero i, Refl))
expr_l (Plus ex ey) = 
  let (x' ** (ex', px)) = expr_l ex in
  let (y' ** (ey', py)) = expr_l ey in
  let (res ** (normRes, Pres)) = plusLExpr ex' ey' in
    (res ** (normRes, rewrite px in (rewrite py in Pres)))
      where 
      plusLExpr : {gam : Vect n Nat} -> {x, y : Nat} 
            -> LExpr gam x -> LExpr gam y  
            -> (z ** (LExpr gam z, x+y=z))
      plusLExpr {x=x} ex LZero =
        (_ ** (ex, rewrite (plusZeroRightNeutral x) in Refl))            
      plusLExpr ex (LPlus e i) =
        let (xRec ** (rec, prfRec)) = plusLExpr ex e in
            (_ ** (LPlus rec i, ?MplusLExpr))

\end{verbatim}
\end{center}
\caption{Production of the left associative form}
\figrule
\end{figure}
In the case of an addition $Plus\ ex\ ey$, the function $expr\_l$ does the job of normalisation recursively on $ex$ and on $ey$, and then it uses the sub-function $plusLExpr$ to normalise the addition of these two --already normalised-- terms. This sub-function $plusLExpr$ has two kind of simplifications to do. When the second argument is an $LZero$, it simply returns its first arguments along with the justification for this rewriting, which obviously uses $plusZeroRightNeutral$. However, when the second argument is an $LPlus\ e\ i$, it continues recursively by computing $plusLExpr\ ex\ e$, and it finally adds $i$ to it. That had for effect to move the parenthesis on the left, and this treatment is going to be justified during the proof of the meta variable $MplusLExpr$ by the use of $plusAssociative$.

This metavariable $MplusLExpr$ requires us to prove the goal : $x1 + (x2 + index\ i\ \Gamma) = xrec + index\ i\ \Gamma$ in a context where we've got, amongst other things, $prfRec : x1 + x2 = xrec$.
By using the property of associativity on the goal, we now need to prove $(x1 + x2) + index\ i\ \Gamma$, which can be done by rewriting the proof $prfRec$ obtained recursively.

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
MplusLExpr = proof {
  intros
  rewrite (sym (plusAssociative x1 x2 (index i gam))); rewrite prf; 
  exact Refl;
}
\end{verbatim}
\end{center}
\caption{Proof of the metavariable plusLExpr1}
\figrule
\end{figure}

It is really important to understand that the root of the automatic building of the desired proof of $x=y$ happens precisely in these nested use of $plusZeroRightNeutral$ and $plusAssociative$ that we've seen in the definition of $expr\_l$ and of its meta-variable $MplusLExpr$. These proofs replace the arithmetical proofs that we were doing previously by hand, like in section 1 with the lemma $adc\_lemma\_2$. \\
\\
Using this new datatype $LExpr$ has changed the representation of our encoded lists, so we need to convert back an $LExpr\ \Gamma\ x$ to an $Expr\ \Gamma\ x$. The function $l\_expr$ does this easy task.
\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
l_expr : LExpr gam x -> Expr gam x
l_expr LZero = Zero
l_expr (LPlus x i) = Plus (l_expr x) (Var i)
\end{verbatim}
\end{center}
\caption{Going back from LExpr to Expr}
\figrule
\end{figure}

We notice that in order to transform the expression into its left associative equivalent representation, we've effectively needed to know where the variables and the $Z$ constants are : the functions $expr\_l$ and $l\_expr$ are doing different treatments for these two possibilities. \\
\\

### 2016-01-13 paper-JFP-DTP/paper.tex
paper continued, section 1&2, typos...

\begin{abstract}
% The abstract should be at most 300 words for the Journal of Functional Programming
Idris is a general purpose purely functional programming language with dependent types, aiming to bring type-based program verification techniques to functional programmers. One common difficulty with programming with dependent types is that proof obligations arise naturally once programs become even moderately sized. For example, implementing an adder for binary numbers indexed over their natural number equivalents will naturally lead to proof obligations for equalities of expressions over natural numbers. 
As far as possible, we would like to solve such proof obligations automatically. In this paper, we show one way to automate such proofs by reflection. We will show how representing Idris expressions in a reflected form, indexed by the original Idris expression, leads to straightforward construction and manipulation of proofs.
With this type-safe reflection technique, the resulting expressions are guaranteed to be faithful representations of the corresponding inputs and any generated proof is guaranteed to be a proof of the required property. 
We will first present the technique on a small example, where we aim to automatically prove equalities with universally quantified natural numbers and the associative addition operation, and later show how this idea can be generalised to various kind of properties (or axioms) that might be available (for example : commutativity, distributivity, existence of neutral elements...), leading to a hierarchy of reflexive tactics for Monoids, Commutative Monoids, Groups, Commutative Groups, Rings, and so on, written in Idris, for proving different kind of equivalence. We will also show how each tactic reuses the other ones from the simplest structures, thus avoiding as much as possible the duplication of code.
\end{abstract}

>>>>

\begin{abstract}
% The abstract should be at most 300 words for the Journal of Functional Programming
Idris is a general purpose purely functional programming language with dependent types, aiming to bring type-based program verification techniques to functional programmers. One common difficulty with programming with dependent types is that proof obligations arise naturally once programs become even moderately sized. For example, implementing an adder for binary numbers indexed over their natural number equivalents will naturally lead to proof obligations for equalities of expressions over natural numbers. 
As far as possible, we would like to solve such proof obligations automatically. In this paper, we show one way to automate such proofs by reflection. We will show how representing Idris expressions in a reflected form, indexed by the original Idris expression, leads to straightforward construction and manipulation of proofs.
With this type-safe reflection technique, the resulting expressions are guaranteed to be faithful representations of the corresponding inputs and any generated proof is guaranteed to be a proof of the required property. 
We will first present the technique on a small example, where we aim to automatically prove equalities with universally quantified natural numbers and the associative addition operation, and will later show how this idea can be generalised to various kind of properties that might be available (for example, commutativity, distributivity, existence of neutral elements...), leading to a hierarchy of reflexive tactics for Monoids, Commutative Monoids, Groups, Commutative Groups, Rings, and so on, written in Idris, for proving different kind of equivalence. We will also show how each tactic reuses the other ones from the simplest structures, thus avoiding as much as possible the duplication of code.
\end{abstract}
